---
title: "climate_data"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

- Prepara climate data 

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(janitor)
library(kableExtra)
library(lubridate)

# Custom function
waterday <- function(x, start_month=10) {
  start_jday <- lubridate::yday(lubridate::ymd(paste(2001, start_month, 1, sep='-'))) - 1
  wday <- lubridate::yday(x)-(start_jday+lubridate::leap_year(x))
  wday <- ifelse(wday<=0, 365+wday+lubridate::leap_year(x), wday)
  wday
}
```



```{r, eval=FALSE}
raw <- read_csv("/Users/ajpelu/Downloads/cn_PSN07_166.csv")
variables <- read.csv("/Users/ajpelu/Downloads/cn_variables.tsv", sep="\t")
dat <- read_csv("/Users/ajpelu/Downloads/PSN07_diario_2021.csv") |> clean_names() 

dat2021 <- dat |> 
  mutate(date = as.Date(format(as.Date(fecha, format = "%d/%m/%Y"),
                        "%Y-%m-%d"))) |> 
           dplyr::select(
    date, 
    tmed = temperatura_media_o_c, 
    tmax = temperatura_maxima_o_c, 
    tmin = temperatura_minimo_o_c, 
    prec = lluvia_acumulada_mm) 


prec_dialy <- raw |> 
  filter(variable_id == 49) |>
  mutate(value = as.numeric(value)) |> 
  mutate(date = as.Date(lubridate::ymd_hms(time), format = "%Y-%m-%d")) |>
  group_by(date) |> 
  summarise(prec = sum(value))

prec_dialy <- prec_dialy |> 
  bind_rows(
    (dat2021 |> 
       dplyr::select(date, prec))
  )

temp_dialy <- raw |> 
  filter(variable_id == 108) |>
  mutate(value = as.numeric(value)) |> 
  mutate(date = as.Date(lubridate::ymd_hms(time), format = "%Y-%m-%d")) |>
  group_by(date) |> 
  summarise(tmin = min(value),
            tmax = max(value),
            tmed = mean(value))


temp_dialy <- temp_dialy |> 
  bind_rows(
    (dat2021 |> 
       dplyr::select(date, tmin, tmax, tmed))
  )

climate <- prec_dialy |> 
  inner_join(temp_dialy)

write_csv(climate, "data/climate_psn07_canar.csv")
```


```{r, message=FALSE}
climate <- read_csv("data/climate_psn07_canar.csv")
```

# Precipitation
## Data from PSN_07 (Estación Meteorológica Cáñar)

```{r}
prec_monthly <- climate |> 
  dplyr::select(date, prec) |> 
  mutate(date_ym = format(date, "%Y-%m")) |> 
  group_by(date_ym) |> 
  summarise(monthly_prec = sum(prec)) |> 
  mutate(date = lubridate::parse_date_time(date_ym, "ym", tz = "CET"))
  
prec_monthly |> 
  ggplot(aes(x=date, y=monthly_prec)) +
  geom_bar(stat = "identity") 
```

```{r}
date_samplings <- as.Date(c("2008-04-29",
                            "2011-04-26",
                            "2014-05-07"),
                            format="%Y-%m-%d")
```

## Cummulative water (hydrological year)

### Cáñar 

```{r}
wyear_canar <- climate |> 
  mutate(
    doy = lubridate::yday(date),
    wday = waterday(date), 
    year = lubridate::year(date),
    month = lubridate::month(date)
  ) |> 
  mutate(wyear = year + month %in% c(10:12)) |> 
  dplyr::select(-tmin, -tmax, -tmed)

wyear_canar_cum <- wyear_canar |> 
  group_by(wyear) |> 
  mutate(cum = cumsum(prec)) 

pre2008canar_w <- wyear_canar_cum  |>
  filter(wyear == 2008) |> 
  filter(date < as.Date("2008-04-29", format="%Y-%m-%d")) |> 
  mutate(ndays = length(cum))
  
pre2011canar_w <- wyear_canar_cum  |>
  filter(wyear == 2011) |> 
  filter(date < as.Date("2011-04-26", format="%Y-%m-%d")) |> 
  mutate(ndays = length(cum))
  
pre2014canar_w <- wyear_canar_cum  |>
  filter(wyear == 2014) |> 
  filter(date < as.Date("2014-05-07", format="%Y-%m-%d"))|> 
  mutate(ndays = length(cum))

prec_points_canar_w <- bind_rows(
  pre2008canar_w |> filter(wday == max(wday)),
  pre2011canar_w |> filter(wday == max(wday)),
  pre2014canar_w |> filter(wday == max(wday)) 
) |> 
  dplyr::select(-prec, -month, -year, -doy) |> 
  mutate(
    station = "Cáñar PSN07 (wyear)"
  )
```

### Arquilla

```{r, message=FALSE}
files <- list.files("data/climate_arquilla/", pattern = "*.csv", full.names = TRUE)

arquilla_aux <- files |> 
  map_dfr(~read_delim(., delim =";")) |> 
  distinct() |> 
  clean_names() |> 
  dplyr::select(-station_id, -station_code, -validation, -variable) 

arquilla <- arquilla_aux |> 
  mutate(value = case_when(
    value < 0 ~ 0,
    TRUE ~ value  
  )) |> 
  mutate(
    doy = lubridate::yday(time),
    wday = waterday(time), 
    year = lubridate::year(time),
    month = lubridate::month(time)
  ) |> 
  filter(time > as.Date("1990-10-01", format = "%Y-%m-%d")) |> 
  mutate(wyear = year + month %in% c(10:12)) |> 
  arrange(time)

rm(arquilla_aux)
```


```{r}
wyear_arquilla_cum <- arquilla |> 
  group_by(wyear) |> 
  mutate(cum = cumsum(value)) 

pre2008arquilla_w <- wyear_arquilla_cum  |>
  filter(wyear == 2008) |> 
  filter(time < as.Date("2008-04-29", format="%Y-%m-%d")) |> 
  mutate(ndays = length(cum))

pre2011arquilla_w <- wyear_arquilla_cum  |>
  filter(wyear == 2011) |> 
  filter(time < as.Date("2011-04-26", format="%Y-%m-%d")) |> 
  mutate(ndays = length(cum))
  
pre2014arquilla_w <- wyear_arquilla_cum  |>
  filter(wyear == 2014) |> 
  filter(time < as.Date("2014-05-07", format="%Y-%m-%d"))|> 
  mutate(ndays = length(cum))

prec_points_arquilla_w <- bind_rows(
  pre2008arquilla_w |> filter(wday == max(wday)),
  pre2011arquilla_w |> filter(wday == max(wday)),
  pre2014arquilla_w |> filter(wday == max(wday)) 
) |> 
  dplyr::select(-value, -month, -year, -doy) |>
  rename(date = time) |> 
  mutate(
    station = "Arquilla 6257I (wyear)"
  )
```

```{r}
wpoints <- bind_rows(
  prec_points_arquilla_w,
  prec_points_canar_w) |> 
  mutate(wyear = as.factor(wyear))
```


```{r}
col <- c("2008" = "lightblue",
         "2011" = "black",
         "2014" = "orange")

wpoints |> 
  ggplot(aes(x=wyear, y=cum, fill=wyear)) +
  geom_bar(stat="identity") +
  scale_fill_manual(values = col, name = "wyear") +
  facet_wrap(~station) + 
  ylab("Cummulative Rainfall (mm)") +
  xlab("Year") +
    theme_bw() + 
  theme(
    panel.grid = element_blank()
  ) 
```


```{r, eval=FALSE, echo=FALSE}
wyear_arquilla_cum |> 
  ggplot(aes(y=cum, x=wday)) + geom_line() +
  facet_wrap(~wyear)

```



```{r}
arquilla_line_avg <- wyear_arquilla_cum |> 
  dplyr::select(wday, cum, wyear) |> 
  group_by(wday) |> 
  summarise(mean = mean(cum, na.rm=TRUE),
          sd = sd(cum, na.rm = TRUE),
          se = sd/sqrt(length(cum))) |> 
  rename(cum = mean) |> 
  mutate(wyear = as.character("1990-2017"))


b <- bind_rows(
  (arquilla_line_avg |> 
     filter(wday < 219)),
  (pre2008arquilla_w |> 
     dplyr::select(wday, cum, wyear) |> 
     mutate(sd = NA, se = NA, wyear = as.character(wyear))),
  (pre2011arquilla_w |> 
     dplyr::select(wday, cum, wyear) |> 
     mutate(sd = NA, se = NA, wyear = as.character(wyear))),
  (pre2014arquilla_w |> 
     dplyr::select(wday, cum, wyear) |> 
     mutate(sd = NA, se = NA, wyear = as.character(wyear))))

plot_lineas_arquilla <- b |> 
  ggplot(aes(x=wday, y=cum, col=as.factor(wyear))) +
  geom_ribbon(aes(ymin = cum - se, ymax = cum + se), 
              fill="gray", alpha = .1, linetype = 2, 
              show.legend = FALSE) + 
  geom_line() +
  theme_bw() + 
  theme(
    panel.grid = element_blank(),
    legend.position = c(.85,.15), 
    legend.background = element_blank()
    
  ) +
  ylab("Cummulative Rainfall (mm), Arquilla 6257I") +
  xlab("Hydrological day") + 
  scale_color_manual(values=c("gray", "lightblue", "black", "orange"), 
                     name = "Hydrological year") +
  geom_point(data = prec_points_arquilla_w, 
             aes(x=wday, y=cum), 
             size = 3) 

plot_lineas_arquilla
```


```{r}
canar_line_avg <- wyear_canar_cum |> 
  dplyr::select(wday, cum, wyear) |> 
  group_by(wday) |> 
  summarise(mean = mean(cum, na.rm=TRUE),
          sd = sd(cum, na.rm = TRUE),
          se = sd/sqrt(length(cum))) |> 
  rename(cum = mean) |> 
  mutate(wyear = as.character("2008-2020"))


bcanar <- bind_rows(
  (canar_line_avg |> filter(wday < 219)),  
  (pre2008canar_w |> 
     dplyr::select(wday, cum, wyear) |> 
     mutate(sd = NA, se = NA, wyear = as.character(wyear))),
  (pre2011canar_w |> 
     dplyr::select(wday, cum, wyear) |> 
     mutate(sd = NA, se = NA, wyear = as.character(wyear))),
  (pre2014canar_w |> 
     dplyr::select(wday, cum, wyear) |> 
     mutate(sd = NA, se = NA, wyear = as.character(wyear))))

bcanar |> 
  ggplot(aes(x=wday, y=cum, col=as.factor(wyear))) +
  geom_ribbon(aes(ymin = cum - se, ymax = cum + se), 
              fill="gray", alpha = .1, linetype = 2, 
              show.legend = FALSE) + 
  geom_line() +
  theme_bw() + 
  theme(
    panel.grid = element_blank()
  ) +
  ylab("Cummulative Rainfall (mm), Cáñar PSN07") +
  xlab("Hydrological day") + 
  scale_color_manual(values=c("lightblue", "gray","black", "orange"), 
                     name = "Hydrological year") +
  geom_point(data = prec_points_canar_w, 
             aes(x=wday, y=cum), 
             size = 3) 
```




### Arquilla previous 30, 60, 90, 120 days
```{r}
pre2008arquilla_ws <- pre2008arquilla_w |> 
  mutate(d2s = (interval(round(time, unit="day"), as.Date("2008-04-29", format = "%Y-%m-%d")) |>
  as.numeric('days'))) |> 
  arrange(d2s) |> 
  mutate(cum_from = cumsum(value)) 
  

pre2011arquilla_ws <- pre2011arquilla_w |> 
  mutate(d2s = (interval(round(time, unit="day"), as.Date("2011-04-26", format = "%Y-%m-%d")) |>
  as.numeric('days'))) |> 
  arrange(d2s) |> 
  mutate(cum_from = cumsum(value)) 
    
pre2014arquilla_ws <- pre2014arquilla_w |> 
  mutate(d2s = (interval(round(time, unit="day"), as.Date("2014-05-07", format = "%Y-%m-%d")) |>
  as.numeric('days'))) |> 
  arrange(d2s) |> 
  mutate(cum_from = cumsum(value)) 
  

d2s <- bind_rows(
  pre2008arquilla_ws, 
  pre2011arquilla_ws,
  pre2014arquilla_ws
) |> 
  filter(d2s %in% c(60, 90, 120, 150))
```

```{r}
d2s |> 
  ggplot(aes(x=as.factor(wyear), y=cum_from, fill=as.factor(wyear))) +
  geom_bar(stat="identity") +
  scale_fill_manual(values = col, name = "wyear") +
  facet_wrap(~d2s) + 
  ylab("Cummulative Rainfall (mm)") +
  xlab("Year") +
    theme_bw() + 
  theme(
    panel.grid = element_blank()
  )
```


### Select 120 days previous to sampling date 

```{r}
d2s120 <- d2s |> 
  filter(d2s == 120) |> 
  ggplot(aes(x=as.factor(wyear), y=cum_from, fill=as.factor(wyear))) +
  geom_bar(stat="identity") +
  scale_fill_manual(values = col, name = "wyear") +
  # facet_wrap(~d2s) + 
  ylab("Cummulative Rainfall (mm)") +
  xlab("") +
    theme_bw() + 
  theme(
    panel.grid = element_blank(), 
    legend.position = "none"
  )
d2s120
```


```{r, fig.height=7, fig.width=8}
library(cowplot)
combined_plot <- ggdraw() + 
  draw_plot(plot_lineas_arquilla) + 
  draw_plot(d2s120, 
            x = -0.21, 
            y = 0.3, 
            scale = 0.32)

combined_plot
```

```{r}
ggsave(plot = combined_plot, 
       filename = here::here("output/cumulative.jpg"), 
       height = 18, width = 18, units = "cm",
       dpi = 600, device = "jpeg")
```



- Cummulative rainfall (bars) in the 120 days prior to the sampling date. Points and error bars indicate the average cumulative value along the period 1990-2017 (i.e. average value of the cummulative precipitations 120 days previous to sampling date every year)

```{r}
from_avg2008 <- wyear_arquilla_cum |> 
  mutate(d2s = (interval(round(time, unit="day"), as.Date(paste0(wyear,"-04-29"), format = "%Y-%m-%d")))
                  |> as.numeric('days')) |> 
  filter(d2s %in% c(0:150)) |> 
  group_by(wyear) |> 
  arrange(d2s) |> 
  mutate(cum_from = cumsum(value)) |> 
  ungroup() |> 
  filter(d2s %in% c(60, 90, 120, 150)) |> 
  group_by(d2s) |> 
  summarise(mean = mean(cum_from, na.rm=TRUE),
          sd = sd(cum_from, na.rm = TRUE),
          se = sd/sqrt(length(cum_from))) |> 
  rename(cum_from = mean) |> 
  mutate(wyear_avg = as.character("1990-2017"), 
         wyear = as.factor(2008))


from_avg2011 <- wyear_arquilla_cum |> 
  mutate(d2s = (interval(round(time, unit="day"), as.Date(paste0(wyear,"-04-26"), format = "%Y-%m-%d")))
                  |> as.numeric('days')) |> 
  filter(d2s %in% c(0:150)) |> 
  group_by(wyear) |> 
  arrange(d2s) |> 
  mutate(cum_from = cumsum(value)) |> 
  ungroup() |> 
  filter(d2s %in% c(60, 90, 120, 150)) |> 
  group_by(d2s) |> 
  summarise(mean = mean(cum_from, na.rm=TRUE),
          sd = sd(cum_from, na.rm = TRUE),
          se = sd/sqrt(length(cum_from))) |> 
  rename(cum_from = mean) |> 
  mutate(wyear_avg = as.character("1990-2017"), 
         wyear = as.factor(2011))
  
from_avg2014 <- wyear_arquilla_cum |> 
  mutate(d2s = (interval(round(time, unit="day"), as.Date(paste0(wyear,"-05-07"), format = "%Y-%m-%d")))
                  |> as.numeric('days')) |> 
  filter(d2s %in% c(0:150)) |> 
  group_by(wyear) |> 
  arrange(d2s) |> 
  mutate(cum_from = cumsum(value)) |> 
  ungroup() |> 
  filter(d2s %in% c(60, 90, 120, 150)) |> 
  group_by(d2s) |> 
  summarise(mean = mean(cum_from, na.rm=TRUE),
          sd = sd(cum_from, na.rm = TRUE),
          se = sd/sqrt(length(cum_from))) |> 
  rename(cum_from = mean) |> 
  mutate(wyear_avg = as.character("1990-2017"), 
         wyear = as.factor(2014))


mean_froms <- bind_rows(
  from_avg2008,
  from_avg2011,
  from_avg2014
)

mean_froms120 <- mean_froms |> filter(d2s == 120) 


d2s |> 
  filter(d2s == 120) |> 
  ggplot(aes(x=as.factor(wyear), y=cum_from)) +
  geom_bar(stat="identity", fill="lightblue", col="lightblue") +
  # scale_fill_manual(values = col, name = "wyear") +
  # facet_wrap(~d2s) + 
  geom_point(
    data = mean_froms120, aes(x=wyear, y=cum_from), col = "black", size = 4,
  ) + 
  geom_errorbar(
    data = mean_froms120, 
    aes(x=wyear, ymin = cum_from - se, ymax= cum_from + se), 
    col = "black",
    width = .15
  ) +
  ylab("Cummulative Rainfall (mm)") +
  xlab("Year") +
    theme_bw() + 
  theme(
    panel.grid = element_blank()
  )


```





```{r, eval=FALSE, echo=FALSE}
pre_cum |> 
  ggplot(aes(x=doy, y=prec)) +
  geom_bar(stat = "identity") +
  facet_wrap(~year, ncol = 1) +
  ylab("Precipitation (mm)") + 
  xlab("Day of Year") +
  theme_bw() + 
  theme(
    panel.grid = element_blank(),
    # panel.grid.major.x = element_blank(),
    strip.background = element_rect(fill=NA), 
    strip.text = element_text(face="italic")
  ) 
  

prec_cum_mean <- climate |> 
  mutate(doy = lubridate::yday(date), 
         year = lubridate::year(date)) |> 
  group_by(year) |> 
  mutate(cum = cumsum(prec)) |> 
  dplyr::select(doy, cum, year) |> 
  group_by(doy) |> 
  summarise(mean = mean(cum, na.rm=TRUE),
          sd = sd(cum, na.rm = TRUE),
          se = sd/sqrt(length(cum))) |> 
  rename(cum = mean) |> 
  mutate(year = as.character("2008-2021"))

prec_cum_line <- 
  pre_cum |> 
  mutate(year = as.factor(year)) |> 
  group_by(year) |> 
  mutate(cum = cumsum(prec)) |> 
  dplyr::select(doy, year, cum) |> 
  mutate(year = as.character(year)) |> 
  relocate(year, .after = doy)

points <- prec_cum_line |> 
 group_by(year) |> 
  filter(doy ==  max(doy))
prec_cum_line_avg <- 
  bind_rows(prec_cum_line, prec_cum_mean) |>
  filter(doy < 127) |> 
  rowwise() |> 
  mutate(days_to_sampling = case_when(
    year == "2008" ~ (doy - 119), 
    year == "2011" ~ (doy - 115), 
    year == "2014" ~ (doy - 126), 
    year == "2008-2021" ~ (doy - 126))) 
    

prec_cum_line_avg |> 
  ggplot(aes(x=doy, y=cum, col=as.factor(year))) +
  geom_ribbon(aes(ymin = cum - se, ymax = cum + se), 
              fill="gray", alpha = .1, linetype = 2, 
              show.legend = FALSE) + 
  geom_line() +
  theme_bw() + 
  theme(
    panel.grid = element_blank()
  ) +
  ylab("Cummulative Rainfall (mm)") +
  xlab("Day of year") + 
  scale_color_manual(values=c("lightblue", "gray", "black", "orange"), 
                     name = "year") +
  geom_point(data = points, 
             aes(x=doy, y=cum), 
             size = 3) 

prec_cum_line <- 
  pre_cum |> 
  group_by(year) |> 
  mutate(cum = cumsum(prec))

points <- prec_cum_line |> 
 group_by(year) |> 
  filter(doy ==  max(doy))

prec_cum_line |> 
  ggplot(aes(x=doy, y=cum, col=as.factor(year))) +
  geom_line() +
  theme_bw() + 
  theme(
    panel.grid = element_blank()
  ) +
  ylab("Cummulative Rainfall (mm)") +
  xlab("Day of year") + 
  scale_color_manual(values=c("lightblue", "black", "orange"), 
                     name = "year") +
  geom_point(data = points, 
             aes(x=doy, y=cum), 
             size = 3)
  

points |> 
  mutate(year = as.factor(year)) |> 
  ggplot(aes(x=year, y=cum, fill=year)) +
  geom_bar(stat="identity") +
    scale_fill_manual(values = col, name = "year") +
  ylab("Cummulative Rainfall (mm), Cáñar PSN07") +
  xlab("Year") +
    theme_bw() + 
  theme(
    panel.grid = element_blank()
  ) 
```


```{r}
wpoints |> 
  dplyr::select(-date) |> 
  rename(`Cummulative precipitation` = cum) |> 
  relocate(station) |> 
  kableExtra::kbl() |> 
  kable_styling()

```


# Temperature 
```{r}
pre2008 <- climate |> 
  filter(date >= as.Date("2008-01-01", format="%Y-%m-%d")) |> 
  filter(date < as.Date("2008-04-29", format="%Y-%m-%d")) 

pre2011 <- climate |>
  filter(date >= as.Date("2011-01-01", format="%Y-%m-%d")) |> 
  filter(date < as.Date("2011-04-26", format="%Y-%m-%d")) 
  
pre2014 <- climate |>
  filter(date >= as.Date("2014-01-01", format="%Y-%m-%d")) |> 
  filter(date < as.Date("2014-05-07", format="%Y-%m-%d"))                     

pre_cum <- 
  bind_rows(
    pre2008, pre2011, pre2014) |> 
  mutate(year = lubridate::year(date), 
         doy = lubridate::yday(date))

t2008 <- pre_cum |> 
  filter(year == 2008) |> 
  filter(date >= as.Date("2008-03-29", format="%Y-%m-%d")) |> 
  arrange(desc(doy)) |> 
  mutate(d = -seq.int(length(doy)))

t2011 <- pre_cum |> 
  filter(year == 2011) |> 
  filter(date >= as.Date("2011-03-26", format="%Y-%m-%d")) |> 
  arrange(desc(doy)) |> 
  mutate(d = -seq.int(length(doy)))


t2014 <- pre_cum |> 
  filter(year == 2014) |> 
  filter(date >= as.Date("2014-04-06", format="%Y-%m-%d"))|> 
  arrange(desc(doy)) |> 
  mutate(d = -seq.int(length(doy))) 

temp <- bind_rows(t2008,t2011,t2014) |> dplyr::select(-prec)

```

```{r}
temp |> 
  mutate(year = as.factor(year)) |> 
  ggplot(aes(x=d, y=tmed, colour=year)) +
  geom_ribbon(aes(ymin = tmin, ymax=tmax, fill=year), 
              colour=NA, alpha=.2) +
  geom_line() +
  scale_fill_manual(values = col, name = "year") +
  scale_colour_manual(values = col, name = "year") +
  ylab("Temperature (º)") +
  xlab("Previous days") +
    theme_bw() + 
  theme(
    panel.grid = element_blank()
  ) 
```

```{r}
temp |> 
  dplyr::select(tmin, tmax, tmed, year) |> 
  group_by(year) |> 
  summarise(across(tmin:tmed, list(
    mean = mean,
    sd = sd))) |> 
  mutate_at(2:7, list(~round(., 2))) |> 
  unite("tmean", starts_with("tmed"), sep = " ± ") |> 
  unite("tmin", starts_with("tmin"), sep = " ± ") |> 
  unite("tmax", starts_with("tmax"), sep = " ± ") |> 
  kableExtra::kbl() |> 
  kable_styling()
```

